### 2014-09-02 standing-variation-paper.tex
changes throughout

@@

recall peromyscus examples with patchy environments

Summarize previous paper

Question assumptions of no standing variation and homogenous selection

\section{General model}

outline here assumptions common to both models.

Stick to constant-speed waves.
Note monty's empirical speeds paper.

\section{Standing variation}

@@

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the haploid fitness
\gc{heterozygous fitness, with no reference to the fitness of the
  homozygote for the mutant alleles}.
The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
(If the mutation is recessive, XXX more complicated things happen.)
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared distance between parent and child is $\sigma^2$ (the {\em dispersal variance}).
The species occupies an area with mean density
$\rho$ haploid individuals (or chromosomes) per unit area.
\marginnote{XXX or do diploid? below is haploid.}

@@

Before $t=0$, on the other hand, the mutation is deleterious, so the
genetic descendants of each new mutation are (with high probability)
doomed (doomed I say!) to extinction,
but may persist for some time.
Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the log of the size of an extant families is small relative to the establishment time,
and the spatial distribution of the family is small relative to the spread between them.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}

\plr{XXX More discussion on whether this is a good approximation?  
Compute typical size and radius of a standing cluster?}
\gc{Point out connection to standard mutation selection balance formula}

\paragraph{Geographic spread of alleles} 
Once an allele has become local established it can begin to spatially spread. 
We assume that the allele, having become established, quickly settles
down to spread as a traveling wave of constant speed. The behavior of
this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}.

If the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance) then the speed of advance
of this wave will be $v = \sigma \sqrt{2s}$.

\plr{recall PPP computations; draw cones of opportunity for new mutations }

\paragraph{}
So we can put these ingredients together for a simple model of the
geographic spread of alleles. A cartoon example of this process is
shown in Figure \ref{fig:cartoon} Initially, when the selection
pressures change at $t=0$, a set of standing variants can start to
spread having escaped loss through drift. These variants are shown as
little lightning bolts, and occur at a density $\lambda_0$ across
space. 
They spread at velocity $v$ carving out cones in space-time. As these
alleles proceed in their geographic spread other new alleles can arise
and become established in parallel, these are shown as little
stars. These new mutations arise and become established at rate $\lambda$.

@@

\subsection{G6PD}

@@

The geographic area of Central and Eastern Asia with malaria is on the
order of $\sim 10$ million $km^2$. In that area there are at least $15$
common variants \citep{Howes-g6pd-variants}. So the average width of
an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km.

G6pd is 515 codons long, with 140 variants reported in it REF. Assuming a
mutation rate of $\approx 10^{-8}$ per base-pair per generation, we assume
$\mu \approx 1 \times 10^{-6}$ a generation. 
The dispersal and demographic parameters of humans is unclear,
particularly as we are concerned with the effective population density.
We use $\rho=2$ and $0.2$ people per $km^2$ as two choices of our effective population density.
We use $\sigma=10,~50,$ and $100$ as our dispersal parameter.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ 
Map of G6pd-deficiency allele frequencies across Asia. The pie chart shows
  the frequency of G6pd-deficiency alleles. The size of the pie chart
  indicates the number of G6pd-deficient individuals sampled.
. Countries with
  endemic malaria are colored yellow. Figure taken from
  \citet{Howes-g6pd-variants} \url{http://www.malariajournal.com/content/12/1/418}.}
\end{center}
\end{figure}

@@

In \citet{ralphcoop2010}, studying the model without standing variation,
we defined a {\em characteristic length} which gave the spatial scale across with mutants with distinct origins would establish.
This was proportional to the mean distance between neighboring established mutants,
but had the advantage of being easier to calculate.
Furthermore, the time scale over which adaptation occurred could be found by dividing the characteristic length 
by the speed at which the mutants spread.
We will define the characteristic length for this new model,
as well as a similar compound parameter describing the relative importance of standing variation to the process of adaptation.
%(Note that a similar approach, summarizing with a {\em characteristic length} was taken by \citep{slatkin1973geneflow} for a different problem.)

@@

\subsection{Standing Variation -- Results}

The quantity that is perhaps easiest to compute is the mean time until adaptation.
Fix some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as shown in Figure XX,
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\]
and so we have
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.

@@

One interesting problem that we can seek to address with this work is
the extent to which

Lets imagine for the moment that all classes of beneficial allele have
the same beneficial selection coefficient. 
Each class of mutations could have its own mutation rate $\mu_j$ and
have a disadvantage $s_{d,j}$ prior to the switch. As they have the
same selection coefficient all of the waves travel outward at a rate
$v$.

\begin{align}
    \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } ; ~~      \lambda_{j} = \mu_j \rho s_b;
\end{align}

\begin{equation}
    z_0 = \int_0^\infty {2 \lambda_{0,j} \pi v^2 t - \lambda_j \pi v^2 t^3 } \; \exp \left(\sum_j - \lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3 \right) dt .
\end{equation}

>>>>

%\iflatexml  % no margin notes in latexml
% \newcommand{\marginnote}[1]{}
%\newcommand{\marginnote}[1]{{\it\color{red}(#1)}}
%\fi

@@

@@

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
  homozygote for the mutant alleles.
The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
(If the mutation is recessive, XXX more complicated things happen.)
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared distance between parent and child is $\sigma^2$ (the {\em dispersal variance}).
The species occupies an area with mean density
$\rho$ haploid individuals (or chromosomes) per unit area.
\marginnote{XXX or do diploid? below is haploid.}

We assume that dispersal occurs by local movement of
individuals. Specifically we will assume that individuals dispersal a
normally distributed distance, with the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance)

@@

Before $t=0$, on the other hand, the mutation is deleterious, so the
genetic descendants of each new mutation are (with high probability)
doomed (doomed I say!) to extinction,
but may persist for some time.
Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the log of the size of an extant families is small relative to the establishment time,
and the spatial distribution of the family is small relative to the spread between them.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}
Note that this is closely related to the standard frequency of a
allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}

\plr{XXX More discussion on whether this is a good approximation?  
Compute typical size and radius of a standing cluster?}

\paragraph{Geographic spread of alleles} 
Once an allele has become local established it can begin to spatially spread. 
We assume that the allele, having become established, quickly settles
down to spread spatially as a traveling wave of constant speed. The behavior of
this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}. Under reasonably general
conditions the speed of advance
of this wave will be $v = \sigma \sqrt{2s}$. See
\citet{ralph2010parallel} for a fuller review of the Fisher-KPP as it
applies to this scenerio. Note that the speed of the wave can be
changed

\paragraph{Putting it together.}
So we can put these ingredients together for a simple model of the
geographic spread of alleles. A cartoon example of this process is
shown in Figure \ref{fig:cartoon} Initially, when the selection
pressures change at $t=0$, a set of standing variants can start to
spread having escaped loss through drift. These variants are shown as
little lightning bolts, and occur at a density $\lambda_0$ across
space. They spread at velocity $v$ carving out cones in space-time. As these
alleles proceed in their geographic spread other new alleles can arise
and become established in parallel, these are shown as little
stars. These new mutations arise and become established at rate $\lambda$.

As we previously outlined in \citet{ralph2010parallel} this model of
geographic convergent evolution, when $\lambda_0=0$, is analogous to the \citet{kolmogorov-crystallization}
model of crystallization. In \citet{kolmogorov-crystallization} model nucleation points form at randomly at
a constant rate in time and space 
and initiate the radial growth of new crystals. The crystals at the
end point of this process form a random tessellation covering space
the properties of which have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
The generalized version of this process, for non-constant wave speeds
and imhomogeneous Poisson processes is termed the 
Kolmogorov-Johnson-Mehl-Avrami tessellation \citep[KJMA, see ][]{fanfoni-tomellini}.
Our combined process with both standing variation and new mutation is
a specially case of the KJMA tessellation, where the
spatial-temporally homogeneous
Poisson origination process of new mutations, is supplemented by a
single pulse of origination points at time zero with spatial density $\lambda_0$.

If we only allow standing variation then everything about the process
is relatively simple. All of the
points in space closer to a particular origination point will be
reached by that allele before any other
allele has a chance to spread there. Formally then the spatial patterning,
given a set of origination points, would be specified by the Voronoi
tessellation (i.e. the cells formed by assigning points to their
nearest origin). All of the properties of the patterning of space by alleles is
determined by the spatial locations of the initiation points, as
specified by the spatially homogeneous Poisson process, independent of
the rate of spatial spread. So that the properties
of the spatial tessellation with standing variation depends only on
$\lambda_0$, and this
model is commonly called the Poisson-Voronoi tesselation.

In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
speeds, as can arise due to fat-tailed dispersal kernels. Here we
restict ourselves to constant wave speeds for simplicity but similar
calculations could be done by expansion of the results of the previous paper.

@@

\subsection{G6PD example}
We will describe a number of properties of our process, but we first
want to introduce a motivating example to provide some concrete
numbers to use for illustrative purposes.

@@

The geographic area of Central and Eastern Asia with malaria is on the
order of $\sim 10$ million $km^2$. In that area there are at least $15$
common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). So the average width of
an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
The coding region of G6pd is 515 codons long, with $\sim$140
functional, distinct variants that cause reduced G6PD activity. Assuming a
mutation rate of $\approx 10^{-8}$ per base-pair per generation, we assume
$\mu \approx 1 \times 10^{-6}$ a generation. 
The dispersal and demographic parameters of humans in the past few
thousans years is unclear,
particularly as we are concerned with the effective population
density. 
We use $\rho=2$ and $0.2$ people per $km^2$ as two choices of our effective population density.
We use $\sigma=10,~50,$ and $100$ as our dispersal parameter. 
We note that obviously human migration has been
shaped both by local dispersal and larger-scale expansions \citep[see
][for a recent discussion]{}, so these parameters are for
illustrative purposes only.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ 
Map of G6pd-deficiency allele frequencies across Asia. The pie chart shows
  the frequency of G6pd-deficiency alleles. The size of the pie chart
  indicates the number of G6pd-deficient individuals sampled.
. Countries with
  endemic malaria are colored yellow. Figure taken from
  \citet{Howes-g6pd-variants}
  \url{http://www.malariajournal.com/content/12/1/418}. \gc{Not sure
    whether to include this.}
} \label{fig-G6PD-map}
\end{center}
\end{figure}

@@

In \citet{ralph2010parallel}, studying the model without standing variation,
we defined a {\em characteristic length} which gave the spatial scale across with mutants with distinct origins would establish.
This was proportional to the mean distance between neighboring established mutants,
but had the advantage of being easier to calculate.
Furthermore, the time scale over which adaptation occurred could be found by dividing the characteristic length 
by the speed at which the mutants spread.
We will define the characteristic length for this new model,
as well as a similar compound parameter describing the relative importance of standing variation to the process of adaptation.
%(Note that a similar approach, summarizing with a {\em characteristic length} was taken by \citep{slatkin1973geneflow} for a different problem.)

@@

\subsection{Standing Variation -- Results} 
The quantity that is perhaps easiest to compute is the mean time until adaptation.
Fix some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as shown in Figure XX,
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\]
and so we have
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.

@@

One interesting problem that we can seek to address with this work is
the extent to which the degree of pleiotropy biases the 
adaptation towards the repeated use of particular subset of loci
(i.e. convergence at genetic level). 
While many alleles may result in the benefical phenotype, many of
these may be unlikely to contribute to adaptation due to their
negative pleiotropic consequences. There are at least two ways that
negative pleiotropy can contribute to high rates of
convergence when adapting when a single change is sufficient for
adaptation. 
First, negative pleiotropic effects can reduce the overall beneficial
selection coefficient of an allele in the new environment making them
unlikely to become established and slow to spread (and in the worst
case making them deleterious). This effect has been studied by a
number of authors \citep{}. A second contribution of negative
pleiotropy to shaping genetic convergence, is that alleles that have
less negative pleiotropy are more likely to segregate before the
environmental shift, and so get a head start in contributing to
adaptation. This effect is seemingly less well studied, perhaps in
part because many modeling approaches tend to use the strong-selection
weak mutation limit and so ignore standing variation. \gc{Is this fair?}

Here we focus primarily on the second effect. 
Lets imagine for the moment that all classes of beneficial allele have
the same beneficial selection coefficient. 
Each class of mutations could have its own mutation rate $\mu_j$ and
have a disadvantage $s_{d,j}$ prior to the switch. As they have the
same selection coefficient all of the waves travel outward at a rate
$v$. 
The the rate of type $k$ standing variants per unit area and the rate of de
novo variants per unit area per generation are
\begin{align}
    \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } ; ~~      \lambda_{j} = \mu_j \rho s_b;
\end{align}
Using these rates it follows that, at the time when every location has been reached by an adaptive allele,
the proportion of the species range covered by alleles of type $k$ is
\begin{equation} \label{eqn-prop-space-allele-k}
    p_k = \int_0^\infty  \big( 2 \lambda_{0,j} \pi v^2 t + \lambda_j
    \pi v^2 t^3 \big)  \; \exp \left(\sum_k - \lambda_{0,k} \pi v^2 t^2 - \lambda_k \pi v^2 t^3 / 3 \right) dt .
\end{equation}
If we only allow standing variation this collapses to 
\begin{equation}  \label{eqn-prop-space-allele-k-standing-room-only}
    p_k = \frac{  \mu_j / \log(1-s_{d,j}) }{\sum_k  \mu_k / \log(1-s_{d,k}) }
\end{equation}
while if we only allow new variation, i.e. if all variants are highly
deleterious before the environment switches, then $p_k =\mu_j / (\sum_k  \mu_k)$

To illustrate some of the properties of this model lets imagine the
somewhat extreme scenerio that there is a single base-pair where
a mutation change is relatively free of negative pleiotropy (class $1$), and a
larger mutational target where changes have more serious pleiotropic
consequences in the ancestral environment (class $2$). We set $s_{d,1}
\leq s_{d,2}$ and $\mu_1=1 \times 10^{-8}$, and think of the second
class mutational target being ten, a hundred, or one thousand base
pairs. We set  . We show the expected proportion of space covered by
the rarer class $1$ mutations in Figure \ref{fig-pleiotropy_calc},
assuming that both classes share a beneficial selection coefficient
$s_b=0.05$ once the environment switches. 
Intuitively the contribution of the rarer mutation decreases as the
mutational target of the other class becomes larger, and as the
difference in the negative pleiotropoc consequences of the two classes
of alleles decreases. The case with standing variation only is the best
case scenerio for the rarer mutation as we force them to have the same
advantage in the new environment, so its rate of introduction is
necessarily lower. However the standing variation only case does seems
to provide a reasonable rule of thumb. Especially for parameter combinations, higher population densities and
high dispersal distances, that increase the contribution of standing
variation (and similarly for high $s_b$).

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{pleiotropy_calc}
\caption{ %
{\bf Proportion of space covered by rarer but less pleiotropic
  mutation}. Empty circles give the result for standing variation only,
  eqn. \eqref{eqn-prop-space-allele-k-standing-room-only}. Lines give
  the result allowing both standing and de novo mutations  eqn. \eqref{eqn-prop-space-allele-k}}   \label{fig-pleiotropy_calc}
\end{center}
\end{figure}
